---
title: "Exploratory Data Analysis of HIV infection dataset"
author: "Denis Torre"
date: "January 24th, 2017"
output: pdf_document
---

``` {r, echo=FALSE}
# Load libraries
source('/Users/denis/Documents/Projects/scripts/Support.R')

# Get input files
infiles <- c('/Users/denis/Documents/Projects/hiv-analysis/hiv-signature-analysis/f2-normalized_expression_data.dir/podocyte_cell_line-vst.txt',
             '/Users/denis/Documents/Projects/hiv-analysis/hiv-signature-analysis/f2-normalized_expression_data.dir/primary_podocyte-vst.txt',
             '/Users/denis/Documents/Projects/hiv-analysis/hiv-signature-analysis/rawdata/hiv_sample_annotation.txt')

# Get dataframes
cellLineExpressionData <- read.table(infiles[1], header=TRUE, row.names='gene_symbol')
primaryExpressionData <- read.table(infiles[2], header=TRUE, row.names='gene_symbol')
sampleAnnotationDataframe <- read.table(infiles[3], header=TRUE, row.names='sample_name')

# Get annotation data
cellLineAnnotationDataframe <- sampleAnnotationDataframe[sampleAnnotationDataframe$cell_type == 'cell_line', c('cell_type','treatment','timepoint','batch')]
primaryAnnotationDataframe <- sampleAnnotationDataframe[sampleAnnotationDataframe$cell_type == 'primary_podocyte', c('cell_type','treatment','patient_group','patient')]
```

# Overview
## Aim
The aim of this report is to perform Exploratory Data Analysis (EDA) on a HIV infection dataset provided by Dr. Weijia Zhang.

## Data
The data was obtained by performing RNA-sequencing on 34 biological samples of podocytes under three experimental conditions: HIV infection, control GFP infection, and untreated.  Raw readcounts were normalized using Variance Stabilizing Transform (VST) from the DESeq2 package.

\ 

The samples can be divided in two groups:

  1. *13 samples* from **patient-derived cell-lines**, divided in three batches.  3 samples are untreated, the other 10 are infected with HIV at different timepoints (6h, 125, 24h, 48h).
  2. *21 samples* from **patient-derived primary podocytes**, collected from 10 different patients.  The samples are divided in HIV infection, control GFP infection and control.
  
\newpage

# PCA Analysis
## Rationale
A PCA analysis was performed to visualize the data, assess the separation of samples from different experimental conditions, and assess presence of batch effects.

## Cell-line Data
The following PCA displays the 13 samples from patient-derived cell lines, colored by batch.  The plot reveals a very significant batch effect, which could be a potential confounding factor for further analyses.

\ 

``` {r, echo=FALSE, fig.height=5, fig.width=5, fig.align='center', dev.args=list(pointsize=10)}
# Run PCA
pcaRes <- runPCA(cellLineExpressionData)

# Get plot matrix
plotMatrix <- pcaRes$x

# Get color legend
colorLegend <- getLegend(rownames(cellLineAnnotationDataframe), cellLineAnnotationDataframe$batch, colors=c('red','blue','black','grey'))

# PCs
PCs <- c('PC1', 'PC2')

# Plot
plot(plotMatrix[,PCs],
     col=colorLegend$colors,
     xlab=pcaRes$varLabels[PCs[1]],
     ylab=pcaRes$varLabels[PCs[2]])

# Legend
legend('topright', legend=names(colorLegend$legend), fill=colorLegend$legend, title='Batch')
```

\newpage

## Primary Podocyte Data

The following PCA displays the 21 samples from patient-derived primary podocytes, colored by patient.  The two samples from patient(s) NK are outliers.

\

``` {r, echo=FALSE, fig.height=4, fig.width=4, fig.align='center', dev.args=list(pointsize=10)}
# Run PCA
pcaRes <- runPCA(primaryExpressionData)

# Get plot matrix
plotMatrix <- pcaRes$x

# Get color legend
colorLegend <- getLegend(rownames(primaryAnnotationDataframe),
                         primaryAnnotationDataframe$patient_group,
                         colors=c('red','blue','black','grey'))

# PCs
PCs <- c('PC1', 'PC2')

# Plot
plot(plotMatrix[,PCs],
     col=colorLegend$colors,
     xlab=pcaRes$varLabels[PCs[1]],
     ylab=pcaRes$varLabels[PCs[2]])

# Legend
legend('topright', legend=names(colorLegend$legend), fill=colorLegend$legend, title='Patient')
```

To better visualize the separation between samples form B patient(s), the PCA was repeated after removing the two samples coming from NK patient(s).  The following plot displays the results of the new analysis, colored by experimental condition.

``` {r, echo=FALSE, fig.height=4, fig.width=4, fig.align='center', dev.args=list(pointsize=9)}
# Run PCA
pcaRes <- runPCA(primaryExpressionData[,grepl('B', colnames(primaryExpressionData))])

# Get plot matrix
plotMatrix <- pcaRes$x

# Get color legend
colorLegend <- getLegend(rownames(primaryAnnotationDataframe),
                         primaryAnnotationDataframe$treatment,
                         colors=c('red','blue','black','grey'))

# PCs
PCs <- c('PC1', 'PC2')

# Plot
plot(plotMatrix[,PCs],
     col=colorLegend$colors,
     xlab=pcaRes$varLabels[PCs[1]],
     ylab=pcaRes$varLabels[PCs[2]])

# Legend
legend('topright', legend=names(colorLegend$legend), fill=colorLegend$legend, title='Treatment')
```

The analysis shows that the samples don't clearly separate by experimental condition, potentially indicating the presence of additional unaccounted confounding factors.